Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-10-10 pkg/disttree/inst/draft/draft.Rnw
some more revision based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2320 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9bb15fcbae5d21f01b86034d67b0086639e10c93

<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
library("disttree")
library("crch")
library("countreg")
library("gamlss")
library("gamlss.cens")
gen.cens(NO, type = "left")
library("RainTyrol")

## HCL palette
pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)
@

@@

## get predicted parameter
# for learndata and testdata
{
  set.seed(7)

@@

  set.seed(7)
  # locally redefine logLik.distforest to use crps in varimp()
  logLik.distforest <- function(object, newdata = NULL, ...) {
    if(is.null(newdata)) {
      newdata <- object$data
    }

    # predict parameter
    pdf <- predict(object, newdata = newdata, type = "parameter")
    df_mu <- pdf$mu
    df_sigma <- pdf$sigma

    # calculate CRPS
    crps <- mean(crps_cnorm(newdata$robs, location = df_mu, scale = df_sigma, lower = 0, upper = Inf), na.rm = TRUE)

    return(structure(crps, df = NA, class = "logLik"))
  }

  vimp_crps <- varimp(Axams_24to4$df, nperm = 1L)

  save(vimp_crps, file = "vimp_crps.rda")

  rm(logLik.distforest) 
}

@@

In weather forecasting, these properties are especially appealing in 
mountainous regions and complex terrain where a wide range of local-scale
effects are not yet resolved by the NWP models. 
Thus, effects with abrupt changes and possibly nonlinear interactions 
might be required to account for site-specific unresolved features.
To illustrate this in practice,
precipitation forecasts are obtained with distributional forests at 95
meteorological stations in a mountainous region in the Alps, covering mainly
Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{fig:all-map}).
More specifically, a Gaussian distribution, left-censored at zero, is employed to model 24-hour total precipitation so that the
zero-censored point mass describes the probability of observing no precipitation 
on a given day (see Figure~\ref{fig:axams-dist}). Forecasts for July are
established based on data from the same month over the years 1985--2012 including
80~covariates derived from a wide range of different NWP quantities. As
Figure~\ref{fig:axams-dist} shows, the station-wise forests yield a full
distributional forecast for each day -- here for one specific day (July~24) at one
station (Axams) over four years (2009--2012) -- based on the previous 24~years
as learning data. The corresponding observations conform reasonably well with
the predictions. In Section~\ref{sec:precipitation} we investigate the
performance of distributional forests in this forecasting task in more detail.
Compared to three alternative zero-censored Gaussian models distributional forests
perform at least on par and sometimes clearly better while requiring no expert
knowledge for the model specification. The three alternatives are: a standard ensemble model
output statistics approach \citep[EMOS,][]{Gneiting+Raftery+Westveld:2005} based
on an NGR, a GAMLSS with regressors prespecified based on meteorological expertise
\citep[following][]{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
\citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
\citep{Messner+Mayr+Zeileis:2017} as an alternative technique for variable selection among all
80~available regressors.

@@

While Breiman and Cutler's random forests \citep{Breiman:2001} grow ensembles of trees
that pick up changes in the location of the response across the covariates,
\emph{distributional forests} employ an ensemble of $T$~\emph{distributional trees}. 
These pick up changes in the ``direction'' of any distribution parameter by
considering the full score vector for choosing splitting variables and split
points. \hl{For each distributional tree a different learning data set is obtained} 
through resampling (bootstrap or subsampling) and in each node only a
random subset of the covariates $\bold{Z}$ is considered. As usual in random
forests, this reduces the correlation among the trees and stabilizes the
variance of the model.

@@

\item \emph{Distributional forest:} All 80~predictor variables are
  considered for learning a forest of 100 trees. Subsampling is employed 
  for each tree using a third of the predictors
  in each split of the tree (\code{mtry}). Parameters are estimated by
  adaptive local likelihood based on the forest weights as described
  in Section~\ref{sec:methods}. 
  \hl{As stopping criteria the minimal number
  of observations to perform a split} (\code{minsplit}) \hl{is set to $50$, 
  the minimal number of observations in a terminal node} (\code{minbucket})
  \hl{is set to $20$ and the significance level for variable selection} (\code{alpha}) 
  \hl{is kept at the default value $1$.}
  % stopping criteria:
  % forest_minsplit = 50, forest_minbucket = 20, forest_mincrit = 0

@@

\begin{figure}[t!]
\centering
\setkeys{Gin}{width=0.7\textwidth}
<<rain_axams_varim, fig=TRUE, echo=FALSE, width=6.5, height=4.2>>=
par(mar = c(2.5,10,1,2))
barplot(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)], 
        horiz = TRUE, las = 1, axes = FALSE,
        xlab = "Variable importance: mean decrease in CRPS",
        font.axis = 3, #list(family="HersheySerif", face=3),
        xlim = c(0,1.1),
        names.arg = gsub("pow", "", names(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)])))
axis(1, at = seq(0,1.6,0.2), las = 1, mgp=c(0,1,0))
@
\caption{\label{fig:axams-varimp} CRPS-based variable importance for the top 10 covariates 
in the distributional forest. Based on data for station Axams, learning period 
1985--2008 and assessed in 2009--2012.}
\end{figure}

@@

After considering only one observational site up to now, a second step
evaluates and compares the competing methods on all 95 available stations. As in the previous
section, all models are learned on the first 24~years and evaluated by the
average CRPS on the last 4~years. More specifically, the CRPS skill score
against the EMOS model is computed for the out-of-sample predictions at each
station and visualized by parallel coordinates plots with boxplots superimposed
in Figure~\ref{fig:all-crps}. Overall, distributional forests have a slightly
higher improvement in CRPSS compared to the two GAMLSS which is best seen
by looking at the boxplots and the green line representing the results
for station Axams. The underlying parallel coordinates additionally
bring out that the prespecified GAMLSS sometimes performs rather differently
(sometimes better, sometimes worse) compared to the two data-driven models.
Values below zero show that, for some stations, EMOS performs better
than the more complex statistical methods.
%NAs in gamlss in 13 stations!!!\\
%for example in 13th station in complete{\_}stations\\
%\text{Error in glim.fit(f = mu.object, X = mu.X, y = y, w = w, fv = mu, os = mu.offset,  :}\\ 
%\text{NA's in the working vector or weights for parameter mu}
%\\
% sum(is.na(crps[,3]))
%<<eval = TRUE, echo = FALSE>>=

@@

>>>>

<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
library("disttree")
library("crch")
library("countreg")
library("gamlss")
library("gamlss.cens")
gen.cens(NO, type = "left")
library("RainTyrol")
library("scoringRules")

## HCL palette
pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)

# set function for parallelization
applyfun <- function(X, FUN, ...) parallel::mclapply(X, FUN, ..., mc.cores = pmax(1, parallel::detectCores() - 1))

@@

## get PIT
{
  set.seed(7)

@@

  set.seed(7)
  nperm <- 50

  # function to permutate chosen variable and then calculate mean crps
  meancrps <- function(permute = NULL, newdata = Axams_24to4$testdata) {
    if(!is.null(permute)) newdata[[permute]] <- sample(newdata[[permute]])
    p <- predict(Axams_24to4$df, newdata = newdata, type = "parameter")
    mean(crps_cnorm(newdata$robs, location = p$mu, scale = p$sigma, lower = 0))
  }

  # apply for all covariates except for dswrf_mean_min and 
  # dswrf_sprd_min (columns 30 and 33) as they are always 0

  # risk <- sapply(c(5:29, 31, 32, 34: ncol(testdata)), meancrps)

  # using only one core
  # risk_all <- replicate(nperm, sapply(c(5:29, 31, 32, 34: ncol(testdata)), meancrps))
  # risk <- rowMeans(risk_all)

  # or parallel
  risklist <- applyfun(1:nperm, 
                       function(i){
                         set.seed(i)
                         sapply(c(5:29, 31, 32, 34: ncol(Axams_24to4$testdata)), meancrps)
                       })
  risk <- Reduce("+", risklist) / length(risklist)

  names(risk) <- names(Axams_24to4$testdata)[c(5:29, 31, 32, 34: ncol(Axams_24to4$testdata))]
  vimp_crps <- risk - meancrps(newdata = Axams_24to4$testdata)
  vimp_crps <- sort(vimp_crps, decreasing = TRUE)

  save(vimp_crps, file = "vimp_crps.rda")
}

@@

In weather forecasting, these properties are especially appealing in 
mountainous regions and complex terrain where a wide range of local-scale
effects are not yet resolved by the NWP models. 
Thus, effects with abrupt changes and possibly nonlinear interactions 
might be required to account for site-specific unresolved features.
To illustrate this in practice,
precipitation forecasts are obtained with distributional forests at 95
meteorological stations in a mountainous region in the Alps, covering mainly
Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{fig:all-map}).
More specifically, a Gaussian distribution, left-censored at zero, is employed to model 24-hour total precipitation so that the
zero-censored point mass describes the probability of observing no precipitation 
on a given day (see Figure~\ref{fig:axams-dist}). Forecasts for July are
established based on data from the same month over the years 1985--2012 including
80~covariates derived from a wide range of different NWP quantities. As
Figure~\ref{fig:axams-dist} shows, the station-wise forests yield a full
distributional forecast for each day -- here for one specific day (July~24) at one
station (Axams) over four years (2009--2012) -- based on the previous 24~years
as learning data. The corresponding observations conform reasonably well with
the predictions. In Section~\ref{sec:precipitation} we investigate the
performance of distributional forests in this forecasting task in more detail.
Compared to three alternative zero-censored Gaussian models distributional forests
perform at least on par and sometimes clearly better while requiring no meteorological
knowledge about the athmospheric processes which drive formation of precipitation
for the model specification. The three alternatives are: a standard ensemble model
output statistics approach \citep[EMOS,][]{Gneiting+Raftery+Westveld:2005} based
on an NGR, a GAMLSS with regressors prespecified based on meteorological expertise
\citep[following][]{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
\citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
\citep{Messner+Mayr+Zeileis:2017} as an alternative technique for variable selection among all
80~available regressors.

@@

\hl{The idea of random forests is to learn an ensemble of trees, each on a different
learning data obtained through resampling (bootstrap or subsampling). In each node 
only a random subset of the covariates $\bold{Z}$ is considered to reduce the 
correlation among the trees and to stabilizes the variance of the model.
For a simple regression random forest the mean of predictions over
all trees is considered. In that way changes in the location  of the response 
across the covariates are detected} (e.g. in Breiman and Cutler's random forests, 
\cite{Breiman:2001}).
\hl{This idea is now taken one step further by embedding it in a distributional 
framework based on loglikelihood estimation. \emph{Distributional forests} 
employ an ensemble of $T$~\emph{distributional trees} which pick up changes in 
the ``direction'' of any distribution parameter by considering the full score 
vector for choosing splitting variables and split points.}

@@

\item \emph{Distributional forest:} All 80~predictor variables are
  considered for learning a forest of 100 trees. Subsampling is employed 
  for each tree using a third of the predictors
  in each split of the tree (\code{mtry}). Parameters are estimated by
  adaptive local likelihood based on the forest weights as described
  in Section~\ref{sec:methods}. 
  \hl{As stopping criteria the minimal number of observations to perform a split} 
  (\code{minsplit}) \hl{is set to $50$, the minimal number of observations in a 
  terminal node} (\code{minbucket})\hl{is set to $20$ and the significance level 
  for variable selection} (\code{alpha}) \hl{is kept at the default value $1$.}
  % stopping criteria:
  % forest_minsplit = 50, forest_minbucket = 20, forest_mincrit = 0

@@

\begin{figure}[t!]
\centering
\setkeys{Gin}{width=0.7\textwidth}
<<rain_axams_varim, fig=TRUE, echo=FALSE, width=6.5, height=4.2>>=
par(mar = c(4,10,1,2))
barplot(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)], 
        horiz = TRUE, las = 1, axes = FALSE,
        xlab = "Variable importance: mean increase in CRPS",
        font.axis = 3, #list(family="HersheySerif", face=3),
        xlim = c(0,0.08),
        names.arg = gsub("pow", "", names(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)])))
axis(1, at = seq(0,1,0.02), las = 1, mgp=c(0,1,0))
@
\caption{\label{fig:axams-varimp} CRPS-based variable importance for the top 10 covariates 
in the distributional forest. Based on data for station Axams, learning period 
1985--2008 and assessed in 2009--2012.}
\end{figure}

@@

After considering only one observational site up to now, a second step
evaluates and compares the competing methods on all 95 available stations. As in the previous
section, all models are learned on the first 24~years and evaluated by the
average CRPS on the last 4~years. More specifically, the CRPS skill score
against the EMOS model is computed for the out-of-sample predictions at each
station and visualized by parallel coordinates plots with boxplots superimposed
in Figure~\ref{fig:all-crps}. Overall, distributional forests have a slightly
higher improvement in CRPSS compared to the two GAMLSS which is best seen
by looking at the boxplots and the green line representing the results
for station Axams. The underlying parallel coordinates additionally
bring out that the prespecified GAMLSS sometimes performs rather differently
(sometimes better, sometimes worse) compared to the two data-driven models.
Values below zero show that, for some stations, EMOS performs better
than the more complex statistical methods.
%NAs in gamlss in 13 stations!!!\\
%for example in 13th station in complete{\_}stations\\
%\text{Error in glim.fit(f = mu.object, X = mu.X, y = y, w = w, fv = mu, os = mu.offset,  :}\\ 
%\text{NA's in the working vector or weights for parameter mu}
%\\
% sum(is.na(crps[,3]))

@@

\subsubsection*{Variable Selection}
\hl{Generally, there are many possibilities how to specify the variables that are
to be included in a model. Especially for a low number of variables all types
of interactions can be employed and mixed effect models and separations into 
main effects and interaction effects can be considered allowing for easier 
interpretations} (see for example \cite{Wood+Scheipl+Faraway:2013}, 
\cite{Goicoa+Adin+Ugarte:2018}, \cite{Ugarte+Adin+Goicoa:2017}).
However, in the application presented in Section~\ref{sec:precipitation}
alltogether 80 variables are available. Even if only paired interactions were
considered this would already lead to 3160 additional covariates in the model
making it very difficult to keep the model manageable regarding computational
costs but also interpretability. Therefore, only main effects are considered
for the distributional forest and the boosted GAMLSS which both automatically
select covariates while meteorological expert knowledge was applied to 
select covariates and specify interactions for the prespecified GAMLSS.
Therefore,

\subsubsection*{Tuning parameters}
Setting tuning parameters of a forest model can have a major influence
not only on the accuracy of the model but also on the computational 
complexity. 
For the application in Section~\ref{sec:precipitation} we have 
selected the values of these parameters (listed in Section~\ref{sec:evaluation}) 
based on advice from literature and our own experience.

\cite{Hastie+Tibshirani+Friedman:2001} and \cite{Breiman:2001} recommend to 
build full-grown trees. Therefore, low values should be chosen for the stopping 
criteria \code{minsplit}, \code{minbucket} and \code{alpha}.
However, in a distributional model it also has to be considered that a 
minimal number of observations is required to allow for a reasonable 
estimation of parameters in each node. For that reason the minimal number
of observations for which statistical tests are performed in the 
distributional forest in Section~\ref{sec:precipitation} is set to
50 (\code{minbucket}) and the minimal number of observations on which
parameters are estimated is set to 20 (\code{minsplit}).

As stated by \cite{Breiman:2001} and \cite{Hastie+Tibshirani+Friedman:2001} 
\hl{random forests do not overfit, not even for an increasing number of 
trees} (\code{ntree}). \hl{Hence, it is recommended to build forests consisting 
of a large number of trees} (\code{ntree}) \hl{as this generally leads to an increase 
in accuracy. However, this also comes along with increasing computational costs. 
Therefore, ``a good choice results from a trade-off between computational 
complexity and accuracy''} \citep{Biau+Scornet:2016}.
\hl{Following this advice we chose to build forests consisting of 100 trees. 
However, increasing this tuning parameter could improve the performance
of the distributional forest without the problem of overfitting.}

%% TO DOs:
%% East Tyrol
\subsubsection*{East Tyrol}
Figure~\ref{fig:all-map} shows that, contrary to the rest of Tyrol, 
for the majority of stations in East Tyrol the prespecified GAMLSS performs
best in this particular setting (learning data: 1985--2008, testing data:
2009--2012). East Tyrol is located south of the main Alpine Ridge and 
exhibits different climatological characteristics than the part north of 
the Alpine Ridge. It seems that the choice of covariates and their interactions 
we made for the prespecified GAMLSS method is particularly adequate for the 
region of East Tyrol for this specific analysis. However, the advantages of 
the prespecified GAMLSS models in this particular setting (training: 1985--2008, 
validation: 2009--2012) even out in a cross-validation framework for most of 
the stations in East Tyrol. These results are presented in the online supplement 
which comes along with our manuscript. There, more extensive evaluations
are carried out for 15 observation stations, including a cross-validation
as presented for station Axams in this paper.

\subsubsection*{Choice of station Axams}
%% choice of station Axams mit Verweis auf Supplement stationwise
Axams was chosen for the more extensive evaluations presented
in this paper as it is the observation station being closest 
to Innsbruck, the capital of Tyrol and the work place of 
three of the authors, but not based on any other particular 
features. Furthermore, it has been checked, whether it 
is an adequate representative of all stations regarding 
the results of our evaluations. This is now also 
confirmed by further analysis in the online supplement which
comes along with our manuscript.
There, the same evaluations as for station Axams have been carried out for 14 other
stations, including some of those where the forest does not perform better than
the other models in the setting as presented in Figure~\ref{fig:all-map}.  
It shows that the differences in performance even out in the cross-validation 
framework. There, the forest performs at least equally well compared to the 
best of the other models at most stations.

\subsubsection{Distributional specifications}
%% choice of distribution mit Verweis auf Supplement distributions
Moreover, as mentioned before we have tested robustness against 
distributional misspecification. The results show that applying a 
left-censored at zero logistic distribution or a Gaussian hurdle model 
instead of the left-censored at zero Gaussian distribution, as presented 
in the paper, does not lead to any significant improvements in goodness of 
fit of the applied models. The exact results are provided in an online 
supplement which comes with our manuscript.

\subsubsection*{Computational difficulties}
%% GAMLSS errors with references gamboostLSS Hofner et al.
As stated by \cite{Hofner+Mayr+Schmid:2016} the AIC-based variable selection 
methods implemented in the \textsf{R} package \textbf{gamlss} ``can be unstable, 
especially when it comes to selecting possibly different sets of variables for 
multiple distribution parameters.'' We have noticed computational problems when
applying the function \textbf{gamlss} in certain settings within the cross-validation
framework as it did not succeed in fitting a model. In this cases the method
was not taken into consideration in the comparison of all applied models.
